---
output:
  html_document:
    theme: paper
    highlight: haddock
---





# Data API Example

```{r setup, include=FALSE}
knitr::opts_chunk$set( echo = TRUE ,  warning = F , message = F )
```

<br>
<br>

### LOAD LIBRARIES

```{r}

library( RCurl )
library( jsonlite )
library( pander )
library( dplyr )

```




### CENSUS API PAGE

http://www.census.gov/data/developers/data-sets.html

<br>
<br>

### REQUEST A KEY

http://api.census.gov/data/key_signup.html

```{r, eval=T}

APIkey <- "your.census.api.key.goes.here"

```


```{r, echo=FALSE}

APIkey <- "358fc9ce0bf0e149d67e9f33cfbb077fa9ca2226"

```

<br>
<br>



### ANATOMY OF AN API


```{r}


fieldnm <- "DP03_0119PE"   # poverty
state <- "36"              # NY
county <- "067"            # Onondaga County
APIkey <-  "your,key"      # http://api.census.gov/data/key_signup.html    

```




<br>
<br>

### FETCH CENSUS DATA

We will retrieve data for the following three variables from Onondaga County in NY.

* poverty rate: DP03_0119PE
* total pop: DP05_0028E
* pop black: DP05_0033E




```{r, echo=F}
json.to.data <- function( x )
{
	a.matrix <- fromJSON(x)  # converts json table to a matrix

	c.names <- a.matrix[ 1 , ]  # column names are the first row

	a.matrix <- a.matrix[ -1 , ]

	my.dat <- data.frame( a.matrix, stringsAsFactors=F )

	names( my.dat ) <- c.names
	
	# my.dat[,1] <- as.numeric( as.character( my.dat[,1] ) )

	# > names( my.dat )
	# [1] "DP03_0119PE" "state"       "county"      "tract"
	
	return( my.dat )
}


poverty.json <- 
'[["NAME","B01001_001E","state"],
["Alabama","4830620","01"],
["Alaska","733375","02"],
["Arizona","6641928","04"],
["Arkansas","2958208","05"],
["California","38421464","06"],
["Colorado","5278906","08"],
["Connecticut","3593222","09"],
["Delaware","926454","10"],
["District of Columbia","647484","11"],
["Florida","19645772","12"],
["Georgia","10006693","13"],
["Hawaii","1406299","15"],
["Idaho","1616547","16"],
["Illinois","12873761","17"],
["Indiana","6568645","18"],
["Iowa","3093526","19"],
["Kansas","2892987","20"],
["Kentucky","4397353","21"],
["Louisiana","4625253","22"],
["Maine","1329100","23"],
["Maryland","5930538","24"],
["Massachusetts","6705586","25"],
["Michigan","9900571","26"],
["Minnesota","5419171","27"],
["Mississippi","2988081","28"],
["Missouri","6045448","29"],
["Montana","1014699","30"],
["Nebraska","1869365","31"],
["Nevada","2798636","32"],
["New Hampshire","1324201","33"],
["New Jersey","8904413","34"],
["New Mexico","2084117","35"],
["New York","19673174","36"],
["North Carolina","9845333","37"],
["North Dakota","721640","38"],
["Ohio","11575977","39"],
["Oklahoma","3849733","40"],
["Oregon","3939233","41"],
["Pennsylvania","12779559","42"],
["Rhode Island","1053661","44"],
["South Carolina","4777576","45"],
["South Dakota","843190","46"],
["Tennessee","6499615","47"],
["Texas","26538614","48"],
["Utah","2903379","49"],
["Vermont","626604","50"],
["Virginia","8256630","51"],
["Washington","6985464","53"],
["West Virginia","1851420","54"],
["Wisconsin","5742117","55"],
["Wyoming","579679","56"],
["Puerto Rico","3583073","72"]]'

```



```{r, results='asis'}



fieldnm <- "DP03_0119PE"   # poverty
state <- "36"              # NY
county <- "067"            # Onondaga County
APIkey <- "your.api.key.here"     

  
paste0(  "http://api.census.gov/data/2013/acs5/profile/?get=", fieldnm,
        "&for=tract:*&in=state:", state,
        "+county:", county,
        "&key=", "your.api.key.here"  )


```


<br>
<br>
    



### FETCH THE DATA

The data is returned in a format that looks like this:

[Census API return](https://api.census.gov/data/2015/acs/acs5?get=NAME,B01001_001E&for=state:*)

<br>
<br>



#### Poverty Rate

```{r, eval=F}

fieldnm <- "DP03_0119PE"   # poverty
state <- "36"              # NY
county <- "067"            # Onondaga County

resURL <-  paste("https://api.census.gov/data/2013/acs/acs5?get=",fieldnm,
               "&for=tract:*&in=state:",state,"+county:",county,"&key=",
               APIkey,sep="")


  
poverty.json <- getURL( resURL, ssl.verifypeer = FALSE )

head( poverty.json )


```


```{r, echo=F}

head( poverty.json )

```

<br>
<br>

That's not very helpful! Let's reformat the json output. 


```{r}

poverty <- json.to.data( poverty.json )

head( poverty ) %>% pander

```


<br>
<br>
<br>
<br>
<br>
<br>


### JSON TO DATA FRAME FUNCTION

```{r, eval=F}
json.to.data <- function( x )
{
	a.matrix <- fromJSON(x)  # converts json table to a matrix

	c.names <- a.matrix[ 1 , ]  # column names are the first row

	a.matrix <- a.matrix[ -1 , ]

	my.dat <- data.frame( a.matrix, stringsAsFactors=F )

	names( my.dat ) <- c.names
	
	# my.dat[,1] <- as.numeric( as.character( my.dat[,1] ) )

	# > names( my.dat )
	# [1] "DP03_0119PE" "state"       "county"      "tract"
	
	return( my.dat )
}

```








